Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2017-11-18 LaTeX/Turnover.tex
fig 1-3 done-ish
compiled up to fig s3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Tighter linkage allows looser sex-linkage to evolve, even in absence of haploid selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter_Mike}}
\caption{
Transitions between XY and ZW systems can occur even when the neo-SDR is less tightly linked to a locus under sexually-antagonistic selection (here, without haploid selection).
In panel A, linkage is loose enough relative to selection that the analytical results assuming weak selection hold, and a neo-W can only invade when it is more tightly linked with the selected locus ($R<r$; shaded region).
In panel B, linkage is tight enough relative to selection that the analytical results assuming weak selection do not hold, and a neo-W can invade even when it is less tightly linked with the selected locus ($r<R$; unshaded region).
In panel C we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$). 
Coloured markers show recombination rates for which the temporal dynamics of invasion are plotted in the inset, demonstrating that neo-W alleles can fix (reach frequency $0.5$ among female gametes) if they are more (black) or less (red) closely linked to a locus experiencing sexually-antagonistic selection. 
A very loosely linked neo-W does not spread in this case (blue and green lines overlap and go to 0). 
Indeed, we can show that neo-W invasion fitness is always negative when $R=1/2$ and there is sex-antagonism but no haploid selection (see supplementary \textit{Mathematica} file). 
Fitness parameters are shown by an asterisk in Figure \ref{fig:regionplots}A: $w_{AA}^\female = 1.05$, $w_{aa}^\male = 1.2$, $w_{aa}^\female = w_{AA}^\male = 0.85$, $w_{Aa}^\Hermaphrodite = 1$,  $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
%\textcolor{red}{consider removing panel A, which is repeated in Figure 3.}
%\textcolor{blue}{probably better to remove from panel 3, eh?}
}
\label{fig:SexAntagTighter}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Sex-ratio selection is not a good predictor of neo-W invasion into an XY system
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Temporal_SR_Mike}
\caption{
Fisherian sex-ratio selection alone is not a good predictor of turnover between sex-determining systems.
In this figure, selection is ploidally antagonistic with haploid selection favouring the $a$ allele during male meiosis.
In panel A, male drive in an ancestral ZW system has no affect on the zygotic sex ratio, yet a neo-Y can invade and replace the ancestral sex-determination system (inset shows neo-Y frequency among male gametes, the ancestral W also goes to fixation during this transition). 
When $R<1/2$, the neo-Y becomes associated with the allele favoured by drive, causing the zygotic sex ratio to become biased, hence the frequency of neo-Y among male gametes can be higher than $0.5$ (inset).
In panel B, male drive in an ancestral XY system causes a male bias, allowing a neo-W to invade and replace the ancestral sex-determination system (inset shows neo-W frequency among female gametes, the ancestral Y also goes to fixation), which balances the zygotic sex ratio.
Parameters:  $s^\female =s^\male = 0.2$, $h^\female = h^\male = 0.7$, $t^\female = t^\male = \alpha^\female_\Delta = 0$, $\alpha^\male_\Delta = -0.1$, $r=0.02$.
}
\label{fig:SexRatioBad}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{PositionPlot_Overdominance_Mike}}
\caption{
Neo-W alleles can spread when loci under diploid selection are tightly linked to the ancestral sex determining locus ($r\approx0$). 
In panels A and B, the $a$ allele is favoured in females ($w_{aa}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{AA}^\female=0.85$) and selection in males is overdominant ($w_{aa}^\male=w_{AA}^\male=0.75$).
In panels C and D, selection in males and females is overdominant ($w_{aa}^\female=w_{AA}^\female=0.6$, $w_{aa}^\male=0.5$, $w_{AA}^\male=0.7$, $w_{Aa}^\Hermaphrodite=1$).
There is no haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
These parameters are marked by daggers in Figure \ref{fig:regionplots}B and C, which show that neo-W invasion is expected for any $R$ ($\lambda_{WA},\lambda_{Wa}>1$) when the $a$ allele is nearly fixed on the Y (black lines in this figure). 
Equilibria where the $A$ allele is more common among Y-bearing male gametes can also be stable and allow neo-W invasion for these parameters (blue lines). 
The weak selection approximation holds when all recombination rates are large relative to selection (around 0 in panels A and C), in which case, in the absence of haploid selection, neo-W alleles should spread only if they are more tightly linked to the selected locus (positive invasion fitness if and only if the selected locus is in the grey region). 
However, when linkage is tight (panels C and D and when the selected locus is near the SDR in all panels), this prediction breaks down. 
}
\label{fig:positionOverdominance}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{Temporal_Overdominance_Mike}}
\caption{
Following invasion by a neo-W allele, there can be a complete transition to a new sex-determination system, maintenance of polymorphism at both ancestral-XY and neo-ZW sex determining regions, or loss of the new sex-determining allele. 
Here we plot the frequency of the neo-W allele among female gametes; as the neo-W reaches frequency $0.5$, polymorphism at the ancestral XY locus is lost with Y becoming fixed such that sex is determined only be the ZW allele carried by a female gamete. 
Panels A, C and D show cases where a steady state is reached with the neo-W at a frequency below $0.5$, in which case ancestral-X and Y alleles also both segregate. 
In all cases, we assume that the $a$ allele is initially more common that the $A$ allele on the Y (Y-$a$ is fixed when $r=0$). 
When $r>0$ (panels B and D), Y-$A$ haplotypes created by recombination can become more common than Y-$a$ haplotypes as the neo-W spreads.
In B, this leads to loss of the neo-W and the system goes to an equilibrium with X-$a$ and Y-$A$ haplotypes fixed (A'), such that all females have the high fitness genotype $aa$ and all males $Aa$. 
For the parameters in B, neo-W alleles have negative invasion fitness when the Y-$A$ haplotype is ancestrally more common than Y-$a$ (see blue line in Figure \ref{fig:temporalOverdominance}A and  \ref{fig:temporalOverdominance}B). 
In contrast, the neo-W is not lost in panel D (see blue line in Figure \ref{fig:temporalOverdominance}C and  \ref{fig:temporalOverdominance}D). 
Fitness parameters are the same as in Figure \ref{fig:temporalOverdominance}, the $a$ allele is favoured in females ($w_{aa}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{AA}^\female=0.85$) and there is overdominant selection in males ($w_{aa}^\male=w_{AA}^\male=0.75$) in panels A and B.
In panels C and D, selection in males and females is overdominant ($w_{aa}^\female=w_{AA}^\female=0.6$, $w_{aa}^\male=0.5$, $w_{AA}^\male=0.7$, $w_{Aa}^\Hermaphrodite=1$). 
These parameters are marked by a dagger in Figure \ref{fig:regionplots}. 
Here, there is no haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
}
\label{fig:temporalOverdominance}
\end{figure}

\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter_MaleDrive_Mike}}
\caption{
When there is sexually-antagonistic selection and haploid selection, a neo-W may invade for any $R$.
Panel A shows that the invasion fitness of a neo-W is positive where linkage is tight, even when $r<R$ (unshaded region). \textcolor{red}{(remove A?)}
In panel B, we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$).
Coloured markers show recombination rates for which the temporal dynamics of neo-W invasion are plotted in panel C (black $R=0.001$, red $R=0.02$, blue $R=0.1$, green $R=0.5$). 
The diploid selection parameters used in this plot are the same as in Figure \ref{fig:SexAntagTighter}, marked by an asterisk in Figure \ref{fig:regionMaleDrive}A: $w_{AA}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{aa}^\female=0.85$, $w_{AA}^\male=0.85$, $w_{aa}^\male=1.05$, $\alpha_{\Delta}^\male=-0.08$, except that there is also male meiotic drive in favour of the $a$ allele, $\alpha_{\Delta}^\male=-0.08$.
When $R=0.5$ (green curve), the neo-W does not reach fixation and X,Y,Z, and W alleles are all maintained in the population, see Figure \ref{fig:freqAll}C.
}
\label{fig:SexAntagTighterMaleDrive}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Tighter linkage allows looser sex-linkage to evolve, even in absence of haploid selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter}}
\caption{
Transitions between XY and ZW systems can occur even when the neo-SDR is less tightly linked to a locus under sexually-antagonistic selection (here, without haploid selection).
In panel A, linkage is loose enough relative to selection that the analytical results assuming weak selection hold, and a neo-W can only invade when it is more tightly linked with the selected locus ($R<r$; shaded region).
In panel B, linkage is tight enough relative to selection that the analytical results assuming weak selection do not hold, and a neo-W can invade even when it is less tightly linked with the selected locus ($r<R$; unshaded region).
In panel C we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$). 
Coloured markers show recombination rates for which the temporal dynamics of invasion are plotted in the inset, demonstrating that neo-W alleles can fix (reach frequency $0.5$ among female gametes) if they are more (black) or less (red) closely linked to a locus experiencing sexually-antagonistic selection. 
A very loosely linked neo-W does not spread in this case (blue and green lines overlap and go to 0). 
Indeed, we can show that neo-W invasion fitness is always negative when $R=1/2$ and there is sex-antagonism but no haploid selection (see supplementary \textit{Mathematica} file). 
Fitness parameters are shown by an asterisk in Figure \ref{fig:regionplots}A: $w_{AA}^\female = 1.05$, $w_{aa}^\male = 1.2$, $w_{aa}^\female = w_{AA}^\male = 0.85$, $w_{Aa}^\Hermaphrodite = 1$,  $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
%\textcolor{red}{consider removing panel A, which is repeated in Figure 3.}
%\textcolor{blue}{probably better to remove from panel 3, eh?}
}
\label{fig:SexAntagTighter}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Sex-ratio selection is not a good predictor of neo-W invasion into an XY system
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Temporal_SR}
\caption{
Fisherian sex-ratio selection alone is not a good predictor of turnover between sex-determining systems.
In this figure, selection is ploidally antagonistic with haploid selection favouring the $a$ allele during male meiosis.
In panel A, male drive in an ancestral ZW system has no affect on the zygotic sex ratio, yet a neo-Y can invade and replace the ancestral sex-determination system (inset shows neo-Y frequency among male gametes, the ancestral W also goes to fixation during this transition). 
When $R<1/2$, the neo-Y becomes associated with the allele favoured by drive, causing the zygotic sex ratio to become biased, hence the frequency of neo-Y among male gametes can be higher than $0.5$ (inset).
In panel B, male drive in an ancestral XY system causes a male bias, allowing a neo-W to invade and replace the ancestral sex-determination system (inset shows neo-W frequency among female gametes, the ancestral Y also goes to fixation), which balances the zygotic sex ratio.
Parameters:  $s^\female =s^\male = 0.2$, $h^\female = h^\male = 0.7$, $t^\female = t^\male = \alpha^\female_\Delta = 0$, $\alpha^\male_\Delta = -0.1$, $r=0.02$.
}
\label{fig:SexRatioBad}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{PositionPlot_Overdominance}}
\caption{
Neo-W alleles can spread when loci under diploid selection are tightly linked to the ancestral sex determining locus ($r\approx0$). 
In panels A and B, the $a$ allele is favoured in females ($w_{aa}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{AA}^\female=0.85$) and selection in males is overdominant ($w_{aa}^\male=w_{AA}^\male=0.75$).
In panels C and D, selection in males and females is overdominant ($w_{aa}^\female=w_{AA}^\female=0.6$, $w_{aa}^\male=0.5$, $w_{AA}^\male=0.7$, $w_{Aa}^\Hermaphrodite=1$).
There is no haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
These parameters are marked by daggers in Figure \ref{fig:regionplots}B and C, which show that neo-W invasion is expected for any $R$ ($\lambda_{WA},\lambda_{Wa}>1$) when the $a$ allele is nearly fixed on the Y (black lines in this figure; not stable for $r>>0$). 
Equilibria where the $A$ allele is more common among Y-bearing male gametes can also be stable and allow neo-W invasion for these parameters (blue lines). 
The weak selection approximation holds when all recombination rates are large relative to selection (around 0 in panels A and C), in which case, in the absence of haploid selection, neo-W alleles should spread if and only if they are more tightly linked to the selected locus (positive invasion fitness if and only if the selected locus is in the grey region). 
However, when linkage is tight (panels B and D and when the selected locus is near the SDRs in all panels), this weak selection prediction can break down. 
}
\label{fig:positionOverdominance}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Both neo-W haplotypes can spread, in the absence of haploid selection, when there is overdominance
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=\linewidth]{Temporal_Overdominance}}
\caption{
Following invasion by a neo-W allele, there can be a complete transition to a new sex-determination system, maintenance of polymorphism at both ancestral-XY and neo-ZW sex determining regions, or loss of the new sex-determining allele. 
Here we plot the frequency of the neo-W allele among female gametes; as the neo-W reaches frequency $0.5$, polymorphism at the ancestral XY locus is lost with Y becoming fixed such that sex is determined only be the ZW allele carried by a female gamete. 
Panels A, C and D show cases where a steady state is reached with the neo-W at a frequency below $0.5$, in which case ancestral-X and Y alleles also both segregate. 
In all cases, we assume that the $a$ allele is initially more common than the $A$ allele on the Y (Y-$a$ is fixed when $r=0$). 
When $r>0$ (panels B and D), Y-$A$ haplotypes created by recombination can become more common than Y-$a$ haplotypes as the neo-W spreads.
In B, this leads to loss of the neo-W and the system goes to an equilibrium with X-$a$ and Y-$A$ haplotypes fixed (equilibrium $A'$), such that all females have the high fitness genotype $aa$ and all males are $Aa$. 
For the parameters in B, neo-W alleles have negative invasion fitness when the Y-$A$ haplotype is ancestrally more common than Y-$a$ (see blue lines in Figure \ref{fig:positionOverdominance}A and  \ref{fig:positionOverdominance}B near the ancestral SDR). 
In contrast, the neo-W is not lost in panel D as it is favoured near $r\approx0$ (see blue lines in Figure \ref{fig:positionOverdominance}C and  \ref{fig:positionOverdominance}D near the ancestral SDR). 
Fitness parameters are the same as in Figure \ref{fig:positionOverdominance}; in panels A and B the $a$ allele is favoured in females ($w_{aa}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{AA}^\female=0.85$) while there is overdominance in males ($w_{aa}^\male=w_{AA}^\male=0.75$) and in panels C and D, there is overdominance in both sexes ($w_{aa}^\female=w_{AA}^\female=0.6$, $w_{aa}^\male=0.5$, $w_{AA}^\male=0.7$, $w_{Aa}^\Hermaphrodite=1$). 
These parameters are marked by a dagger in Figure \ref{fig:regionplots}. 
Here, there is no haploid selection $t^\Hermaphrodite = \alpha^\Hermaphrodite_\Delta = 0$.
}
\label{fig:temporalOverdominance}
\end{figure}

\begin{figure}[!h]
\centering
\centerline{\includegraphics[width=1.5\linewidth]{PositionPlot_SexAntagTighter_MaleDrive}}
\caption{
When there is sexually-antagonistic selection and haploid selection, a neo-W may invade for any $R$.
Panel A shows that the invasion fitness of a neo-W is positive where linkage is tight, even when $r<R$ (unshaded region). \textcolor{red}{(remove A?)}
In panel B, we vary the recombination rate between the neo-W and the selected locus ($R$) for a fixed recombination rate between the ancestral-SDR and the selected locus ($r=0.005$).
Coloured markers show recombination rates for which the temporal dynamics of neo-W invasion are plotted in panel C (black $R=0.001$, red $R=0.02$, blue $R=0.1$, green $R=0.5$). 
The diploid selection parameters used in this plot are the same as in Figure \ref{fig:SexAntagTighter}, marked by an asterisk in Figure \ref{fig:regionMaleDrive}A: $w_{AA}^\female=1.05$, $w_{Aa}^\Hermaphrodite=1$, $w_{aa}^\female=0.85$, $w_{AA}^\male=0.85$, $w_{aa}^\male=1.05$, $\alpha_{\Delta}^\male=-0.08$, except that there is also male meiotic drive in favour of the $a$ allele, $\alpha_{\Delta}^\male=-0.08$.
When $R=0.5$ (green curve), the neo-W does not reach fixation and X,Y,Z, and W alleles are all maintained in the population, see Figure \ref{fig:freqAll}C.
}
\label{fig:SexAntagTighterMaleDrive}
\end{figure}
